\pkg{plotROC}: A Better Tool for Plotting ROC Curves

### 2015-03-13 Paper/sachs2015plotROC.tex
response to first editorial review

\documentclass[article]{jss}

@@

\Shorttitle{\pkg{plotROC}: A Better Tool for Plotting ROC Curves} %% a short title (if necessary)
%% an abstract and keywords
\Abstract{
  Plots of the receiver operating characteristic (ROC) curve are
  ubiquitous in medical research. Designed to simultaneously display the
  operating characteristics at every possible value of a continuous
  diagnostic test, ROC curves are used in oncology to evaluate screening,
  diagnostic, prognostic and predictive biomarkers. I reviewed a sample of
  ROC curve plots from the major oncology journals in order to assess
  current trends in usage and design elements. My review suggests that ROC
  curve plots are often ineffective as statistical charts and that poor
  design obscures the relevant information the chart is intended to
  display. I describe my new \proglang{R} package that was created to
  address the shortcomings of existing tools. The package has functions to
  create informative ROC curve plots, with sensible defaults, for use in
  print or as an interactive web-based plot. A web application was
  developed to reach a broader audience of scientists who do not use R.
}
\Keywords{ROC curves; graphics; interactive; plots}
\Plainkeywords{ROC curves; graphics; interactive; plots} %% without formatting
%% at least one keyword must be supplied

@@

\subsection{About ROC Curves}\label{about-roc-curves}

@@

\subsection{Design Considerations}\label{design-considerations}

@@

\subsection{Existing Plotting
Software}\label{existing-plotting-software}

The ROC curve plot is, at the most basic level, a line graph. Therefore,
once the appropriate statistics are estimated, existing plotting
functions can be used to create an ROC curve plot. Viewers can identify
ROC plots through context, by observing the shape of the line, and
through the addition of axis labels, titles, legends, and so on. In my
review of the oncology literature, I observed plots with the distinctive
characteristics of the plotting functions from Microsoft Office,
\proglang{SAS}, \proglang{SPSS}, and the base \proglang{R} plotting
functions \citep{arr}.

There are several \proglang{R} packages related to ROC curve estimation
that contain dedicated plotting functions. The \pkg{ROCR} package
\citep{rocr} plots the FPF \emph{versus} TPF, as usual, and then takes
the interesting approach of encoding the cutoff values as a separate
color scale along the ROC curve itself. A legend for the color scale is
placed along the vertical axis on the right of the plotting region. The
\pkg{pROC} package \citep{pROC} is mainly focused on estimating
confidence intervals and regions for restricted ranges of the ROC curve.
The plotting methods therein use the base \proglang{R} plotting
functions to create nice displays of the curves along with shaded
confidence regions. My \pkg{plotROC} package uses the \pkg{ggplot2}
\citep{ggplot2} plotting library to create clear, informative ROC plots,
with interactive features for use on the web, and sensible defaults for
use in print.

@@

Each image consisted of a single ROC curve plot or a panel of multiple
plots. Each plot was inspected manually for the following design
features: the number of curves displayed, the type of axis labels
(sensitivity/ 1 - specificity or true/false positive fractions),
presence or absence of grid lines, presence or absence of a diagonal
guide line, whether any cutpoints were indicated, the type of curve
label (legend or direct label), and presence of other textual
annotations such as the AUC. The numerical results of the survey are
summarized in table \ref{table1}.

\begin{table}[ht]
\centering
\begin{tabular}{ll}
  \hline
 & percent (count) \\ 
  \hline
Number of curves &  \\ 
  $\quad$1 & 19.6 (9) \\ 
  $\quad$2 & 43.5 (20) \\ 
  $\quad$3 & 10.9 (5) \\ 
  $\quad$4+ & 26.1 (12) \\ 
  $\quad$Average (SD) & 2.6 (1.5) \\ 
  Axis labels &  \\ 
  $\quad$FPF/TPF & 13.0 (6) \\ 
  $\quad$mixed & 2.2 (1) \\ 
  $\quad$none & 2.2 (1) \\ 
  $\quad$sens/spec & 82.6 (38) \\ 
  Diagonal Guide & 43.5 (20) \\ 
  Gridlines & 17.4 (8) \\ 
  Cutoffs indicated & 15.2 (7) \\ 
  AUC indicated & 50.0 (23) \\ 
  Curve Labels &  \\ 
  $\quad$direct & 10.9 (5) \\ 
  $\quad$legend & 63.0 (29) \\ 
  $\quad$none & 19.6 (9) \\ 
  $\quad$title & 6.5 (3) \\ 
   \hline
\end{tabular}
\caption{Results of a literature review of major oncology journals for ROC curve plots. The rows indicate the frequency and count of key design elements of an ROC curve plot. FPR = False positive rate; TPR = True positive rate; sens = Sensitivity; spec = Specificity; AUC = Area under the Curve} 
\label{table1}
\end{table}

@@

The panels of figure \ref{figure1} illustrate the most common styles of
ROC curve plots, and the associated design elements. I favor the use of
gridlines and a diagonal reference line to facilitate accurate readings
off of the axes. Direct labels are preferred over legends because they
omit the additional cognitive step of matching line types or colors to
labels. My \pkg{plotROC} package additionally provides plotting of
cutoff values, which are displayed interactively with the web-based
output option, and direct labels for print use. Exact confidence regions
for points on the ROC curve are optionally calculated and displayed.
Additionally, I use axis scales adjusted to be denser near the margins 0
and 1. In medical applications, it is often necessary to have a very low
FPR (less than 10\%, for instance), therefore the smaller scales are
useful for accurately determining values near the margins. The next
section details the usage of the \pkg{plotROC} \proglang{R} package and
these features.

\begin{figure}[htbp]
\centering
\includegraphics{figure/figure1-1.pdf}
\caption{Illustration of design choices in plotting ROC curves. Panel A
shows a sparse ROC curve, with no design additions inside the plotting
region. The plot results in more white space than anything else. It is
difficult to accurately determine values without reference lines. Panel
B shows a plot comparing 2 curves, with different line types and a
legend. AUCs are also given in the legend. Panels B and C add gridlines,
diagonal reference lines, and direct labels. \label{figure1}}
\end{figure}

\section{Usage of the Package}\label{usage-of-the-package}

@@

\begin{verbatim}
install.packages("plotROC")
library(plotROC)
\end{verbatim}

@@

\begin{verbatim}
shiny_plotROC()
\end{verbatim}

@@

I start by creating an example data set. The marker I generate is
moderately accurate for predicting disease status.

\begin{verbatim}
D.ex <- rbinom(100, size = 1, prob = .5)
M.ex <- rnorm(100, mean = D.ex)
\end{verbatim}

Next I use the \texttt{calculate\_roc} function to compute the empirical
ROC curve. The disease status need not be coded as 0/1, but if it is
not, \texttt{calculate\_roc} assumes (with a warning) that the lowest
value in sort order signifies disease-free status. This returns a
dataframe with three columns: the cutoff values, the TPF and the FPF.

\begin{verbatim}
roc.estimate <- calculate_roc(M.ex, D.ex)
str(roc.estimate)
\end{verbatim}

\begin{verbatim}
'data.frame':   100 obs. of  3 variables:
 $ c  : num  -2.18 -1.67 -1.57 -1.55 -1.54 ...
 $ TPF: num  1 1 1 1 1 1 1 1 1 1 ...
 $ FPF: num  0.981 0.962 0.943 0.925 0.906 ...
\end{verbatim}

The \texttt{rocdata} is passed to the \texttt{ggroc} function with an
optional label. This creates a ggplot object of the ROC curve using the
\pkg{ggplot2} package \citep{ggplot2}.

\begin{verbatim}
single.rocplot <- ggroc(roc.estimate, label = "Example")
\end{verbatim}

The \texttt{myrocplot} object can be used to create an interactive plot
and display it in the Rstudio viewer or default web browser by passing
it to the \texttt{plot\_interactive\_roc} function. Give the function an
optional path to an html file as an argument called \texttt{file} to
save the interactive plot as a complete web page. A screen shot of an
interactive plot is shown in figure \ref{interact}. Hovering over the
display shows the cutoff value at the point nearest to the cursor.
Clicking makes the cutoff label stick until the next click, and if
confidence regions are available, clicks will also display those as grey
rectangles.

\begin{verbatim}
plot_interactive_roc(single.rocplot)
\end{verbatim}

@@

An interactive ROC plot can be exported by using the
\texttt{export\_interactive\_roc} function, which returns a character
string containing the necessary \proglang{HTML} and
\proglang{JavaScript}. The character string can be copy-pasted into an
html document, or better yet, incorporated directly into a dynamic
document using \pkg{knitr} \citep{knitr}. In a \pkg{knitr} document, it
is necessary to use the \texttt{cat} function on the results and use the
chunk options
\texttt{results = \textquotesingle{}asis\textquotesingle{}} and
\texttt{fig.keep = \textquotesingle{}none\textquotesingle{}} so that the
interactive plot is displayed correctly. For documents that contain
multiple interactive plots, it is necessary to assign each plot a unique
name using the \texttt{prefix} argument of
\texttt{export\_interactive\_roc}. This is necessary to ensure that the
\proglang{JavaScript} code manipulates the correct svg elements. For
examples of interactive plots and how to incorporate them into
\pkg{knitr} documents, see the package vignette
(\texttt{vignette("examples", package = "plotROC")}) or the web page
https://sachsmc.github.io/plotROC/. The next code block shows an example
\pkg{knitr} chunk that can be used in an .Rmd document to display an
interactive plot.

\begin{verbatim}
```{r int-no, results = 'asis', fig.keep = 'none'}
cat(
  export_interactive_roc(single.rocplot, 
                        prefix = "a")
  )
```
\end{verbatim}

The same \texttt{ggroc} object that I called \texttt{myrocplot} can be
used to generate an ROC plot suitable for use in print. It annotates the
cutoff values and is completely in black and white. A simple example
with the default options is shown in figure \ref{first}.

\begin{verbatim}
plot_journal_roc(single.rocplot)
\end{verbatim}

\begin{figure}[htbp]
\centering
\includegraphics{figure/print-1.pdf}
\caption{Illustration of ROC curve plot generated by \pkg{plotROC} for
use in print. \label{first}}
\end{figure}

@@

If you have multiple tests of different types measured on the same
subjects, you can use the \texttt{calculate\_multi\_roc} function to
compute the empirical ROC curve for each test. It returns a list of data
frames with the estimates and cutoff values. Then the
\texttt{multi\_ggroc} function creates the appropriate type of
\texttt{ggplot} object. Confidence regions are not supported for
multiple curves at the time of writing.

\begin{verbatim}
D.ex <- rbinom(100, 1, .5)

paired.data <- data.frame(M1 = rnorm(100, mean = D.ex), 
                       M2 = rnorm(100, mean = D.ex, sd = .4), 
                       M3 = runif(100), D = D.ex)

estimate.list <- calculate_multi_roc(paired.data, c("M1", "M2", "M3"), "D")
\end{verbatim}

Labels can be added easily with the \texttt{label} option of
\texttt{multi\_ggroc}. The length of the label element should match the
number of plotted curves. The \texttt{multi\_ggroc} object can be passed
to the \texttt{plot\_journal\_roc} and the
\texttt{export\_interactive\_roc} functions, as desired. The resulting
plot is shown in figure \ref{multi}.

\begin{verbatim}
multi.rocplot <- multi_ggroc(estimate.list, label = c("M1", "M2", "M3"))
plot_journal_roc(multi.rocplot)
\end{verbatim}

\begin{figure}[htbp]
\centering
\includegraphics{figure/multi2-1.pdf}
\caption{Illustration of \pkg{plotROC} plot with multiple curves.
\label{multi}}
\end{figure}

Both \texttt{plot\_journal\_roc} and \texttt{export\_interactive\_roc}
support a number of options to customize the look of the plots. By
default, multiple curves are distinguished by different line types and
direct labels. For multiple ROC curves that are similar to one another,
those defaults can make it difficult to interpret the plots. Therefore,
we also support colors and legends. The x- and y-axis can be changed by
passing options to \texttt{ggroc} or \texttt{multi\_ggroc}. The next
code block illustrates the available options.

\begin{verbatim}
colorplot <- multi_ggroc(estimate.list, 
                         xlabel = "1 - Specificity", 
                         ylabel = "Sensitivity")
cat(
  export_interactive_roc(colorplot, lty = rep(1, 3), 
                         color = c("black", "purple", "orange"), 
                         legend = TRUE)
  )
\end{verbatim}

@@

I use the \texttt{ci = TRUE} option in \texttt{calculcate\_roc} and
\texttt{ggroc} to compute confidence regions for points on the ROC curve
using the \citet{clopper1934use} exact method. Briefly, exact confidence
intervals are calculated for the \(FPF\) and \(TPF\) separately, each at
level \(1 - \sqrt{1 - \alpha}\). Based on result 2.4 from
\citet{pepe2003statistical}, the cross-product of these intervals yields
a \(100 * (1 - \alpha)\) percent rectangular confidence region for the
pair. The significance level can be specified using the \texttt{alpha}
option.

\begin{verbatim}
roc.ci <- calculate_roc(paired.data$M1, paired.data$D, ci = TRUE, alpha = 0.05)
ci.rocplot <- ggroc(roc.ci, label = "CI Example", ci = TRUE)
\end{verbatim}

@@

\begin{verbatim}
cat(
  export_interactive_roc(ci.rocplot, 
                         prefix = "aci")
  )
\end{verbatim}

For use in print, I pass a small vector of cutoff locations at which to
display the confidence regions. This is shown in figure \ref{conf}.

\begin{verbatim}
plot_journal_roc(ci.rocplot, n.cuts = 10, 
                 ci.at = c(-.5, .5, 2.1))
\end{verbatim}

\begin{figure}[htbp]
\centering
\includegraphics{figure/printci-1.pdf}
\caption{Illustration of \pkg{plotROC} plot with exact confidence
regions. \label{conf}}
\end{figure}

@@

\pkg{plotROC} uses the \pkg{ggplot2} package to create the objects to be
plotted. Therefore, themes and annotations can be added in the usual
\pkg{ggplot2} way. A \texttt{plot\_journal\_roc} figure with a new
theme, title, axis label, and AUC annotation is shown in figure
\ref{annotate}.

\begin{verbatim}
library(ggplot2)
plot_journal_roc(ci.rocplot, n.cuts = 10, 
                 ci.at = c(-.5, .5, 2.1)) + 
  theme_grey() + 
  geom_abline(intercept = 0, slope = 1, color = "white") + 
  ggtitle("Themes and annotations") + 
  annotate("text", x = .75, y = .25, 
           label = "AUC = 0.80") +
  scale_x_continuous("1 - Specificity", breaks = seq(0, 1, by = .1))
\end{verbatim}

\begin{figure}[htbp]
\centering
\includegraphics{figure/print2-1.pdf}
\caption{Using \pkg{ggplot2} themes and annotations with \pkg{plotROC}
objects. \label{annotate}}
\end{figure}

@@

By default \texttt{calculate\_roc} computes the empirical ROC curve.
There are other estimation methods out there, as I have summarized in
the introduction. Any estimation method can be used, as long as the
cutoff, the TPF and the FPF are returned. Then you can simply pass those
values in a data frame to the \texttt{ggroc} function. For example, let
us use the binormal method to create a smooth curve. This approach
assumes that the test distribution is normal conditional on disease
status.

\begin{verbatim}
D.ex <- rbinom(100, 1, .5)
M.ex <- rnorm(100, mean = D.ex, sd = .5)

mu1 <- mean(M.ex[D.ex == 1])
mu0 <- mean(M.ex[D.ex == 0])
s1 <- sd(M.ex[D.ex == 1])
s0 <- sd(M.ex[D.ex == 0])
c.ex <- seq(min(M.ex), max(M.ex), length.out = 300)

binorm.roc <- data.frame(c = c.ex, 
                             FPF = pnorm((mu0 - c.ex)/s0), 
                             TPF = pnorm((mu1 - c.ex)/s1)
                             )
\end{verbatim}

Then I can pass this data.frame to the \texttt{ggroc} function as
before. The example is shown in figure \ref{binorm}.

\begin{verbatim}
binorm.plot <- ggroc(binorm.roc, label = "Binormal")
plot_journal_roc(binorm.plot)
\end{verbatim}

\begin{figure}[htbp]
\centering
\includegraphics{figure/binormal-1.pdf}
\caption{Illustration of smooth binormal ROC curve. \label{binorm}}
\end{figure}

Another potential use of this approach is for plotting time-dependent
ROC curves for time-to-event outcomes estimated as desribed in
\citep{heagerty2000time}.

\section{How it Works}\label{how-it-works}

\pkg{plotROC} makes use of \pkg{ggplot2} \citep{ggplot2}, \pkg{gridSVG}
\citep{gridsvg}, and \pkg{d3.js} \citep{bostock2011d3} to create
interactive plots. The first step in the process is to create
\texttt{ggplot} objects using the \texttt{ggroc} or
\texttt{multi\_ggroc} functions. These functions return standard
\texttt{ggplot} objects that include basic styling, hidden cutoff
labels, and hidden confidence regions. They can be plotted and inspected
in the \proglang{R} console. These form the basis for both the print
versions and the interactive versions of the plots. Creating a print
version by using the \texttt{plot\_journal\_roc} function simply makes
visible a subset of the hidden cutoff labels and confidence regions, if
available.

\pkg{plotROC} makes interactive plots by first converting the
\texttt{ggplot} object into a scalable vector graphic (svg) object with
the \texttt{grid.export} function. This function maps each element of
the plot to a corresponding element of the svg markup language. I keep
track of the names of the points and labels elements so that I can add
interactivity using \pkg{d3.js} and \proglang{JavaScript}. The main
interactive feature I wanted was to be able to display the cutoff labels
at the points on the ROC curve closest to the mouse cursor.

>>>>

\documentclass[codesnippet]{jss}

@@

\Shorttitle{\pkg{plotROC}: A Better Tool for Plotting ROC Curves} %% a short title (if necessary)
%% an abstract and keywords
\Abstract{
  Plots of the receiver operating characteristic (ROC) curve are
  ubiquitous in medical research. Designed to simultaneously display the
  operating characteristics at every possible value of a continuous
  diagnostic test, ROC curves are used in oncology to evaluate screening,
  diagnostic, prognostic and predictive biomarkers. I reviewed a sample of
  ROC curve plots from the major oncology journals in order to assess
  current trends in usage and design elements. My review suggests that ROC
  curve plots are often ineffective as statistical charts and that poor
  design obscures the relevant information the chart is intended to
  display. I describe my new \proglang{R} package that was created to
  address the shortcomings of existing tools. The package has functions to
  create informative ROC curve plots, with sensible defaults, for use in
  print or as an interactive web-based plot. A web application was
  developed to reach a broader audience of scientists who do not use
  \proglang{R}.
}
\Keywords{ROC curves; graphics; interactive; plots}
\Plainkeywords{ROC curves; graphics; interactive; plots} %% without formatting
%% at least one keyword must be supplied

@@

\subsection{About ROC curves}\label{about-roc-curves}

@@

\subsection{Design considerations}\label{design-considerations}

@@

\subsection{Existing plotting
software}\label{existing-plotting-software}

The ROC curve plot is, at the most basic level, a line graph. Therefore,
once the appropriate statistics are estimated, existing plotting
functions can be used to create an ROC curve plot. Viewers can identify
ROC plots through context, by observing the shape of the line, and
through the addition of axis labels, titles, legends, and so on. In my
review of the oncology literature, I observed plots with the distinctive
characteristics of the plotting functions from Microsoft Excel
\citep{excel}, \proglang{SAS} \citep{sas}, \proglang{SPSS} \citep{spss},
and the base \proglang{R} plotting functions \citep{arr}.

There are several \proglang{R} packages related to ROC curve estimation
that contain dedicated plotting functions. The \pkg{ROCR} package
\citep{rocr} plots the FPF \emph{versus} TPF, as usual, and then takes
the interesting approach of encoding the cutoff values as a separate
color scale along the ROC curve itself. A legend for the color scale is
placed along the vertical axis on the right of the plotting region. Due
to the popularity of \pkg{ROCR}, my package supports the performance
objects as returned by \pkg{ROCR}. The \pkg{pROC} package \citep{pROC}
is mainly focused on estimating confidence intervals and regions for
restricted ranges of the ROC curve. The plotting methods therein use the
base \proglang{R} plotting functions to create nice displays of the
curves along with shaded confidence regions. My \pkg{plotROC} package
uses the \pkg{ggplot2} \citep{ggplot2} plotting library to create clear,
informative ROC plots, with interactive features for use on the web, and
sensible defaults for use in print.

@@

Each image consisted of a single ROC curve plot or a panel of multiple
plots. Each plot was inspected manually for the following design
features: the number of curves displayed, the type of axis labels
(sensitivity/ 1 - specificity or true/false positive fractions),
presence or absence of grid lines, presence or absence of a diagonal
guide line, whether any cutpoints were indicated, the type of curve
label (legend or direct label), and presence of other textual
annotations such as the AUC. The numerical results of the survey are
summarized in Table \ref{table1}.

\begin{table}[ht]
\centering
\begin{tabular}{ll}
  \hline
 & Percent (count) \\ 
  \hline
Number of curves &  \\ 
  $\quad$1 & 19.6 (9) \\ 
  $\quad$2 & 43.5 (20) \\ 
  $\quad$3 & 10.9 (5) \\ 
  $\quad$4+ & 26.1 (12) \\ 
  $\quad$Average (SD) & 2.6 (1.5) \\ 
  Axis labels &  \\ 
  $\quad$FPF/TPF & 13.0 (6) \\ 
  $\quad$mixed & 2.2 (1) \\ 
  $\quad$none & 2.2 (1) \\ 
  $\quad$sens/spec & 82.6 (38) \\ 
  Diagonal Guide & 43.5 (20) \\ 
  Gridlines & 17.4 (8) \\ 
  Cutoffs indicated & 15.2 (7) \\ 
  AUC indicated & 50.0 (23) \\ 
  Curve Labels &  \\ 
  $\quad$direct & 10.9 (5) \\ 
  $\quad$legend & 63.0 (29) \\ 
  $\quad$none & 19.6 (9) \\ 
  $\quad$title & 6.5 (3) \\ 
   \hline
\end{tabular}
\caption{Results of a literature review of major oncology journals for ROC curve plots. The rows indicate the frequency and count of key design elements of an ROC curve plot. FPR = False positive rate; TPR = True positive rate; sens = Sensitivity; spec = Specificity; AUC = Area under the Curve} 
\label{table1}
\end{table}

@@

The panels of Figure \ref{figure1} illustrate the most common styles of
ROC curve plots, and the associated design elements. I favor the use of
gridlines and a diagonal reference line to facilitate accurate readings
off of the axes. Direct labels are preferred over legends because they
omit the additional cognitive step of matching line types or colors to
labels. My \pkg{plotROC} package additionally provides plotting of
cutoff values, which are displayed interactively with the web-based
output option, and direct labels for print use. Exact confidence regions
for points on the ROC curve are optionally calculated and displayed.
Additionally, I use axis scales adjusted to be denser near the margins 0
and 1. In medical applications, it is often necessary to have a very low
FPR (less than 10\%, for instance), therefore the smaller scales are
useful for accurately determining values near the margins. The next
section details the usage of the \pkg{plotROC} \proglang{R} package and
these features.

\begin{Schunk}
\begin{figure}
\includegraphics{figure/figure1-1} \caption[Illustration of design choices in plotting ROC curves]{Illustration of design choices in plotting ROC curves. Panel A shows a sparse ROC curve, with no design additions inside the plotting region. The plot results in more white space than anything else. It is difficult to accurately determine values without reference lines. Panel B shows a plot comparing 2 curves, with different line types and a legend. AUCs are also given in the legend. Panels C and D add gridlines, diagonal reference lines, and direct labels. \label{figure1}}\label{fig:figure1}
\end{figure}
\end{Schunk}

\section{Usage of the package}\label{usage-of-the-package}

@@

\begin{Schunk}
\begin{Sinput}
R> install.packages("plotROC")
R> library(plotROC)
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> shiny_plotROC()
\end{Sinput}
\end{Schunk}

@@

I start by creating an example data set. The marker I generate is
moderately accurate for predicting disease status. Next I use the
\code{calculate_roc} function to compute the empirical ROC curve. The
disease status need not be coded as 0/1, but if it is not,
\code{calculate_roc} assumes (with a warning) that the lowest value in
sort order signifies disease-free status. This returns a dataframe with
three columns: the cutoff values, the TPF and the FPF.

\begin{Schunk}
\begin{Sinput}
R> D.ex <- rbinom(100, size = 1, prob = .5)
R> M.ex <- rnorm(100, mean = D.ex)
R> roc.estimate <- calculate_roc(M.ex, D.ex)
R> head(roc.estimate)
\end{Sinput}
\begin{Soutput}
          c TPF       FPF
1 -2.182331   1 0.9811321
2 -1.674012   1 0.9622642
3 -1.568181   1 0.9433962
4 -1.550585   1 0.9245283
5 -1.539929   1 0.9056604
6 -1.190555   1 0.8867925
\end{Soutput}
\end{Schunk}

The \code{roc.estimate} object is passed to the \code{ggroc} function
with an optional label. This creates a \code{ggplot} object of the ROC
curve using the \pkg{ggplot2} package \citep{ggplot2}.

\begin{Schunk}
\begin{Sinput}
R> single.rocplot <- ggroc(roc.estimate, label = "Example")
\end{Sinput}
\end{Schunk}

The \code{single.rocplot} object can be used to create an interactive
plot and display it in the Rstudio viewer or default web browser by
passing it to the \code{plot_interactive_roc} function. Give the
function an optional path to an html file as an argument called
\code{file} to save the interactive plot as a complete web page. A
screen shot of an interactive plot is shown in Figure \ref{interact}.
Hovering over the display shows the cutoff value at the point nearest to
the cursor. Clicking makes the cutoff label stick until the next click,
and if confidence regions are available, clicks will also display those
as grey rectangles.

\begin{Schunk}
\begin{Sinput}
R> plot_interactive_roc(single.rocplot)
\end{Sinput}
\end{Schunk}

@@

An interactive ROC plot can be exported by using the
\code{export_interactive_roc} function, which returns a character string
containing the necessary \proglang{HTML} and \proglang{JavaScript}. The
character string can be copy-pasted into an html document, or better
yet, incorporated directly into a dynamic document using \pkg{knitr}
\citep{knitr}. In a \pkg{knitr} document, it is necessary to use the
\code{cat} function on the results and use the chunk options
\code{results = 'asis'} and \code{fig.keep = 'none'} so that the
interactive plot is displayed correctly. For documents that contain
multiple interactive plots, it is necessary to assign each plot a unique
name using the \code{prefix} argument of \code{export_interactive_roc}.
This is necessary to ensure that the \proglang{JavaScript} code
manipulates the correct svg elements. For examples of interactive plots
and how to incorporate them into \pkg{knitr} documents, see the package
vignette (\code{vignette("examples", package = "plotROC")}) or the web
page https://sachsmc.github.io/plotROC/. The next code block shows an
example \pkg{knitr} chunk that can be used in an .Rmd document to
display an interactive plot.

\begin{Code}
```{r int-no, results = 'asis', fig.keep = 'none'}
cat(
  export_interactive_roc(single.rocplot, 
                        prefix = "a")
  )
```
\end{Code}

The same \code{ggroc} object that I called \code{single.rocplot} can be
used to generate an ROC plot suitable for use in print. It annotates the
cutoff values and is completely in black and white. A simple example
with the default options is shown in Figure \ref{first}.

\begin{Schunk}
\begin{Sinput}
R> plot_journal_roc(single.rocplot)
\end{Sinput}
\begin{figure}
\includegraphics{figure/print-1} \caption{Illustration of ROC curve plot generated by \pkg{plotROC} for use in print. \label{first}}\label{fig:print}
\end{figure}
\end{Schunk}

@@

If you have multiple tests of different types measured on the same
subjects, you can use the \code{calculate_multi_roc} function to compute
the empirical ROC curve for each test. It returns a list of data frames
with the estimates and cutoff values. Then the \code{multi_ggroc}
function creates the appropriate type of \code{ggplot} object.
Confidence regions are not supported for multiple curves at the time of
writing.

\begin{Schunk}
\begin{Sinput}
R> D.ex <- rbinom(100, 1, .5)
R> 
R> paired.data <- data.frame(M1 = rnorm(100, mean = D.ex), 
+                        M2 = rnorm(100, mean = D.ex, sd = .4), 
+                        M3 = runif(100), D = D.ex)
R> 
R> estimate.list <- calculate_multi_roc(paired.data, 
+                                      c("M1", "M2", "M3"), "D")
\end{Sinput}
\end{Schunk}

Labels can be added easily with the \code{label} option of
\code{multi_ggroc}. The length of the label element should match the
number of plotted curves. The \code{multi_ggroc} object can be passed to
the \code{plot_journal_roc} and the \code{export_interactive_roc}
functions, as desired. The resulting plot is shown in Figure
\ref{multi}.

\begin{Schunk}
\begin{Sinput}
R> multi.rocplot <- multi_ggroc(estimate.list, label = c("M1", "M2", "M3"))
R> plot_journal_roc(multi.rocplot)
\end{Sinput}
\begin{figure}
\includegraphics{figure/multi2-1} \caption{Illustration of \pkg{plotROC} plot with multiple curves. \label{multi} }\label{fig:multi2}
\end{figure}
\end{Schunk}

Both \code{plot_journal_roc} and \code{export_interactive_roc} support a
number of options to customize the look of the plots. By default,
multiple curves are distinguished by different line types and direct
labels. For multiple ROC curves that are similar to one another, those
defaults can make it difficult to interpret the plots. Therefore, we
also support colors and legends. The x- and y-axis can be changed by
passing options to \code{ggroc} or \code{multi_ggroc}. The next code
block illustrates the available options.

\begin{Schunk}
\begin{Sinput}
R> colorplot <- multi_ggroc(estimate.list, 
+                          xlabel = "1 - Specificity", 
+                          ylabel = "Sensitivity")
R> cat(
+   export_interactive_roc(colorplot, lty = rep(1, 3), 
+                          color = c("black", "purple", "orange"), 
+                          legend = TRUE)
+   )
\end{Sinput}
\end{Schunk}

@@

I use the \code{ci = TRUE} option in \code{calculcate_roc} and
\code{ggroc} to compute confidence regions for points on the ROC curve
using the \citet{clopper1934use} exact method. Briefly, exact confidence
intervals are calculated for the \(FPF\) and \(TPF\) separately, each at
level \(1 - \sqrt{1 - \alpha}\). Based on result 2.4 from
\citet{pepe2003statistical}, the cross-product of these intervals yields
a \(100 * (1 - \alpha)\) percent rectangular confidence region for the
pair. The significance level can be specified using the \code{alpha}
option.

\begin{Schunk}
\begin{Sinput}
R> roc.ci <- calculate_roc(paired.data$M1, paired.data$D, 
+                         ci = TRUE, alpha = 0.05)
R> ci.rocplot <- ggroc(roc.ci, label = "CI Example", ci = TRUE)
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> cat(
+   export_interactive_roc(ci.rocplot, 
+                          prefix = "aci")
+   )
\end{Sinput}
\end{Schunk}

For use in print, I pass a small vector of cutoff locations at which to
display the confidence regions. This is shown in Figure \ref{conf}.

\begin{Schunk}
\begin{Sinput}
R> plot_journal_roc(ci.rocplot, n.cuts = 10, 
+                  ci.at = c(-.5, .5, 2.1))
\end{Sinput}
\begin{figure}
\includegraphics{figure/printci-1} \caption{Illustration of \pkg{plotROC} plot with exact confidence regions. \label{conf}}\label{fig:printci}
\end{figure}
\end{Schunk}

@@

\pkg{plotROC} uses the \pkg{ggplot2} package to create the objects to be
plotted. Therefore, themes and annotations can be added in the usual
\pkg{ggplot2} way. A \code{plot_journal_roc} figure with a new theme,
title, axis label, and AUC annotation is shown in Figure \ref{annotate}.

\begin{Schunk}
\begin{Sinput}
R> library(ggplot2)
R> plot_journal_roc(ci.rocplot, n.cuts = 10, 
+                  ci.at = c(-.5, .5, 2.1)) + 
+   theme_grey() + 
+   geom_abline(intercept = 0, slope = 1, color = "white") + 
+   ggtitle("Themes and annotations") + 
+   annotate("text", x = .75, y = .25, 
+            label = "AUC = 0.80") +
+   scale_x_continuous("1 - Specificity", breaks = seq(0, 1, by = .1))
\end{Sinput}
\begin{figure}
\includegraphics{figure/print2-1} \caption{Using \pkg{ggplot2} themes and annotations with \pkg{plotROC} objects. \label{annotate}}\label{fig:print2}
\end{figure}
\end{Schunk}

@@

By default \code{calculate_roc} computes the empirical ROC curve. There
are other estimation methods out there, as I have summarized in the
introduction. Any estimation method can be used, as long as the cutoff,
the TPF and the FPF are returned. Then you can simply pass those values
in a data frame to the \code{ggroc} function. For example, let us use
the binormal method to create a smooth curve. This approach assumes that
the test distribution is normal conditional on disease status.

\begin{Schunk}
\begin{Sinput}
R> D.ex <- rbinom(100, 1, .5)
R> M.ex <- rnorm(100, mean = D.ex, sd = .5)
R> 
R> mu1 <- mean(M.ex[D.ex == 1])
R> mu0 <- mean(M.ex[D.ex == 0])
R> s1 <- sd(M.ex[D.ex == 1])
R> s0 <- sd(M.ex[D.ex == 0])
R> c.ex <- seq(min(M.ex), max(M.ex), length.out = 300)
R> 
R> binorm.roc <- data.frame(c = c.ex, 
+                              FPF = pnorm((mu0 - c.ex)/s0), 
+                              TPF = pnorm((mu1 - c.ex)/s1)
+                              )
\end{Sinput}
\end{Schunk}

Then I can pass this \code{data.frame} to the \code{ggroc} function as
before. The example is shown in Figure \ref{binorm}.

\begin{Schunk}
\begin{Sinput}
R> binorm.plot <- ggroc(binorm.roc, label = "Binormal")
R> plot_journal_roc(binorm.plot)
\end{Sinput}
\begin{figure}
\includegraphics{figure/binormal-1} \caption[Illustration of smooth binormal ROC curve]{Illustration of smooth binormal ROC curve. \label{binorm}}\label{fig:binormal}
\end{figure}
\end{Schunk}

\pkg{plotROC} also supports the use of the \pkg{ROCR} estimation method.
Objects as returned by the function \code{rocr::performance} can be
passed directly to \code{ggroc}:

\begin{Schunk}
\begin{Sinput}
R> library(ROCR)
R> rocr.est <- performance(prediction(M.ex, D.ex), "tpr", "fpr")
R> rocr.plot <- ggroc(rocr.est, label = "ROCR object")
\end{Sinput}
\end{Schunk}

Another potential use of this approach is for plotting time-dependent
ROC curves for time-to-event outcomes estimated as desribed in
\citep{heagerty2000time}. Here is an example using the \pkg{survivalROC}
package \citep{survroc} for estimation:

\begin{Schunk}
\begin{Sinput}
R> library(survivalROC)
R> survT <- rexp(350, 1/5)
R> cens <- rbinom(350, 1, .1)
R> M <- -8 * sqrt(survT) + rnorm(350, sd = survT)
R> sroc <- lapply(c(2, 5, 10), function(t){ 
+   stroc <- survivalROC(Stime = survT, status = cens, marker = M, 
+                        predict.time = t, method = "NNE", 
+                        span = .25 * 350^(-.2))
+   data.frame(TPF = stroc[["TP"]], FPF = stroc[["FP"]], 
+              c = stroc[["cut.values"]], 
+              time = rep(stroc[["predict.time"]], 
+                         length(stroc[["FP"]])))
+   })
R> 
R> surv.plot <- multi_ggroc(sroc, label = paste("t =", c(2, 5, 10)))
\end{Sinput}
\end{Schunk}

\section{How it works}\label{how-it-works}
